---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data to be proceesed is originally in file "activity.csv", which is assumed 
to be part of the current working directory. The data is read into a dataframe 
named "activity". For that, the following R code is used: 

```{r reading data, echo = TRUE}
activity <- read.csv("activity.csv")
```

The following are some initial exploratory operations to better understand the data. 

### Dimensions of the data are: 
```{r dimensions of data}
dimsactivity <- dim(activity)
print(paste("Number of rows =", dimsactivity[1]))
print(paste("Number of columns =", dimsactivity[2]))
```
Determine the number of observations per day. 
```{r}
library(dplyr)
gactivityDate <- group_by(activity, date)
dataPerDay <- dplyr::summarise(gactivityDate, count = n())
nobs <- unique(dataPerDay$count)
if(length(nobs) == 1) {
      print(paste("There are ", nobs[1], " observations for each day in the data set."))
      n = 24*60/5
      if(nobs[1]==n)
            print("This is exactly the number of 5-minute intervals in 24 hours.")
      else 
            print(paste("The number of 5-minute intervals in 24 hours is", n, "."))
} else {
      print("Observations per day differ. Some of those numbers are:")
      print(nobs)
}
```
This number correspo
2. A new column is added with values 1..7, idendifying the day of the week 
that each registered date corresponds to. The name of the column is _wdayNumber_
and is of type _integer_.
```{r adding column for days}
library(lubridate)
activity$wdayNumber <- day(activity$date)

```

## What is mean total number of steps taken per day?
```{r}
library(dplyr)
gactivityDate <- group_by(activity, date)
sactivityDate <- dplyr::summarise(gactivityDate, totalSteps = sum(steps, na.rm = TRUE))
hist(sactivityDate$totalSteps, breaks = 25, col = "gray", 
     main="Distribution of Total Steps Taken on Each Day", 
     ylab = "Frequency (Number of days)", xlab = "Total Number of Steps Per Day")
md <- round(median(sactivityDate$totalSteps), 2)
mn <- round(mean(sactivityDate$totalSteps), 2)
abline(v=md, col = "red", lwd=2)
abline(v=mn, col="blue", lwd=2)
#title(main="Total Number of Steps Per Day Recorded")
legend("topright", lty=1, lwd=2, legend=c(paste("median = ", md), 
                                     paste("mean = ", mn)), 
                                     col = c("red", "blue"))

```

The following might alse be useful for this analysis. 
```{r} 
library(dplyr)
gactivityDate <- group_by(activity, date)
sactivityDate <- dplyr::summarise(gactivityDate, totalSteps = sum(steps, na.rm = TRUE))
sactivityDate$n <- 1:nrow(sactivityDate)

barplot(sactivityDate$totalSteps, names.arg = sactivityDate$n, 
        xlab = "Day Recorded", ylab = "Total Number of Steps")
md <- round(median(sactivityDate$totalSteps), 2)
mn <- round(mean(sactivityDate$totalSteps), 2)
abline(h=md, col = "red")
abline(h=mn, col="blue")
title(main="Total Number of Steps Per Day Recorded")
legend("top", lty=1, lwd=2, legend=c(paste("median = ", md), 
                                     paste("mean = ", mn)), 
                                     col = c("red", "blue"))
```
... add beef describing what is shown...


## What is the average daily activity pattern?
To explore the average daily activity, we want to determine the average number 
of steps taken on each 5-minute interval. This can be done as follows. 
```{r}
library(dplyr)
gactivityInterval <- group_by(activity, interval)
sactivityInterval <- 
      dplyr::summarise(gactivityInterval, 
                       averageSteps = mean(steps, na.rm = TRUE))
sactivityInterval$intervalNumber <- 1:nrow(sactivityInterval)
plot(sactivityInterval$intervalNumber, sactivityInterval$averageSteps, type = "l",
     xlab = "5-minute Interval in a Day", ylab = "Average Number of Steps in Interval")
maxSteps <- max(sactivityInterval$averageSteps)
maxStepsPeriod <- subset(sactivityInterval, averageSteps == maxSteps)
print(paste("The 5-minute period with maximum average number of steps is: interval", 
            maxStepsPeriod$intervalNumber))
h <- maxStepsPeriod$interval
print(paste("It corresponds to the hour: ", 
            hm(paste(trunc(h/100), h%%100, sep = ":"))))

md <- round(median(sactivityInterval$averageSteps), 2)
mn <- round(mean(sactivityInterval$averageSteps), 2)
abline(h=md, col = "red")
abline(h=mn, col="blue")
title(main="Average Number of Steps Per 5-minute Interval Across All Days")
legend("topright", lty=1, lwd=2, legend=c(paste("median = ", md), 
                                     paste("mean = ", mn)), 
                                     col = c("red", "blue"))

```
... add beef describing what is shown...


## Imputing missing values

We can use the following loop to count how many rows contain na values for each 
variable in the data frame.
```{r}
nm <- names(activity)
for(v in 1:length(nm)) { 
      nmv <- sum(is.na(activity[, v]))  
      if (nmv > 0) print(paste("Variable", nm[v], "has ", nmv, "missing values."))
}
```
We can see that only variable "steps" has missing values. In total, there are
2304 missing values of that variable. 

To assign a value to each of those observations having missing value for variable
steps, what we will do is the following. To each 5-minute period with na value, 
assign the current average number
of steps taken across all days in the data set in which the particular interval 
is not na. However, since there seems to be a clear difference in the activity
during weekend and week days, for those missing values in weekend days, we will
use the corresponding average for those days only, and do the anologous to those
corresponding to weekdays. In those cases (if any) in which all the values are
missing, we shall assign value 0. 

...

## Are there differences in activity patterns between weekdays and weekends?
Now let's add a new column to the activity data frame. This is a column of type 
factor and two levels: weekend and weekday. It classifies each of the 
observations in data frame activity depending to the day of the week that its
date corresponds to. 
```{r adding new factor attribute for days: dayType }
library(lubridate)
weekendDays <- activity$wdayNumber == 1 | activity$wdayNumber == 7
activity$dayType <- "weekday"
activity[weekendDays,]$dayType = "weekend"
activity$dayType <- as.factor(activity$dayType)
```
And verifying how many of the observations belongs to each of the two categories, 
we get the following: 
```{r count observations in each of the two categories} 
table(activity$dayType)
```

Now, let's plot to compare patterns in weekdays vs weekends. 


```{r} 
library(lattice)
library(dplyr)
gactivityDaytypeInterval <- group_by(activity, dayType, interval)
sactivityDaytypeInterval <- dplyr::summarise(gactivityDaytypeInterval, average=mean(steps, na.rm = TRUE))
sactivityDaytypeInterval$intervalNumber <- rep(1:(nrow(sactivityDaytypeInterval)/2), 2)
#rangey <- range(sactivityDaytypeInterval$average)
with(sactivityDaytypeInterval, 
     xyplot(average~intervalNumber|dayType, type = "l",
            xlab = "5-minute Interval", 
            ylab = "Average Number of Steps",
            layout=c(1,2)) 
     )

```